Impacts on population indices if scientific surveys are excluded from marine protected areas

### 2023-12-05 text/main.tex
Revisions

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1.15in]{geometry}
\usepackage{amsmath, amsfonts, amssymb}
\usepackage[dvipsnames]{xcolor}
\definecolor{niceblue}{HTML}{236899}
\newcommand{\rev}[1]{{\color{niceblue} #1}}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=black,
    filecolor=black,
    urlcolor=niceblue,
    citecolor=niceblue,
    linkbordercolor = white
}
\usepackage{longtable}
\usepackage{amsbsy}
\usepackage{epsfig}
\usepackage{bm}
\usepackage{xspace}
\usepackage{color}
\usepackage{lineno}
\usepackage{ragged2e}
\usepackage{comment}

@@

% \doublespacing
\onehalfspacing
\begin{document}
% \begin{spacing}{1.2}
\maketitle

@@

% TODO ICES: 200 words max
% TODO: perhaps could frame this as broader than MPAs (e.g., mostly wind farms in the US) and emphasize the general approach more with the NSB MPA case study
Marine protected areas (MPAs) are increasingly implemented worldwide and typically limit commercial and recreational fishing activities.
However, scientific surveys that impact benthic habitat are sometimes also limited or excluded.
We examine the impacts of excluding scientific surveys on population indices using a proposed MPA network off western Canada, four surveys, and \nSpp\ groundfish species as a case study.
Survey exclusion removed 10--25\% of the area, excluded up to $\approx$50\% of species density, and resulted in minor impacts for most species but considerable impacts for some.
Less precise indices, indices for species more represented in MPAs, and indices for species whose density moved into or out of MPAs were most impacted.
We quantify expected losses of precision, inter-annual accuracy, and trend bias for a percent increase in species density covered by MPAs.
We find that redistributing survey effort outside MPAs would lessen precision loss but with limited benefits to accuracy or trend detection.
While these changes may not qualitatively alter stock assessment outcomes for most species, this is a critical next question.
Losses of power to detect 50\% population declines of up to $\approx$30 percentage points for some species in the most affected survey suggest the impacts may be meaningful in some cases.
If survey restrictions continue expanding, index integrity would further degrade, eventually compromising the ability to manage exploited populations.
Regulating surveys within MPA boundaries therefore requires careful consideration to balance MPA objectives with the need for reliable data.

% to precision, inter-annual accur
% indices with up to \maxPrecisionLost \% reductions in precision, accuracy loss (median absolute relative error) of up to \maxMARE\, and trends deviating by up to $\sim$40\% per decade.

% ---the Northern Shelf Bioregion Marine Protected Area Network---
% would have impacted the precision, accuracy, and trend bias of population indices for \nSpp\ benthic fish species.

% For species with at least 15\% of their density within the MPAs, precision reduced by \precisionWCHGshrunk \% in the most impacted trawl survey and \precisionHBLLshrunk \% in a hook-and-line survey, on average.
% Precision reduced by \covPrecisionGeo\ per 1\% of species density excluded by the survey.
% The coefficient of variation, median absolute relative error increased by \covMareGeo,

% Maximum word count for abstracts: 200 words

% Longer draft:
%% Scientific surveys provide critical data for the assessment of population status, including through indices of population biomass or abundance through time.
%% Marine protected areas (MPAs) are increasingly implemented as one of several approaches to meet spatial conservation targets.
%% While MPAs typically limit commercial or recreational activities, scientific surveys that cause impacts to MPA objectives such as benthic habitat conservation may also be restricted.
%% The impacts of such exclusion to the calculation of indices of abundance across a suite of species and survey types is not well quantified.
%% Here, we conduct a retrospective analysis asking to what extent a proposed network of MPAs---the Northern Shelf Bioregion Marine Protected Area Network off the coast of British Columbia, Canada---would have impacted the precision (coefficient to variation; CV), accuracy (median absolute relative error; MARE), and bias of population indices (relative error trend) of \nSpp\ benthic fish species from three bottom trawl and one hook-and-line depth-stratified randomized surveys.
%% Our results show that the spatial restrictions, which range from 10\% to 30\% across surveys, result in impacts to indices ranging from relatively minor to considerable with up to 40\% increases in imprecision, 40\% increases in inaccuracy, and trends in relative error of up to 40\% per decade.
%% For species with at least 15\% of their density within the MPA, indices of abundance would on average increase their CV 25\% in the most impacted trawl survey and 15\% in a hook-and-line survey.
%% Indices that were less precise to start with suffered larger impacts to accuracy.
%% Indices for species with a higher density proportion in MPAs suffered larger losses of precision, accuracy, and trend bias.
%% Species with a changing proportion of density within vs. outside the MPA suffered the most extreme trend biases as the species became more present or hidden from the survey.
%% We further conclude that reduced survey effort overall is responsible for most of the precision loss, but redistributing survey effort outside the MPA would do little to reduce accuracy loss and trend bias.
%% We also find that the spatially contiguous pattern of MPA restrictions, or the particular MPA locations themselves, are more impactful to accuracy and especially trend bias than an equivalent random effort reduction.
%% Our results are generally similar for geostatistical model-based or design-based estimators.
%%
%% Together, our results suggest small to moderate impacts on survey index integrity for most species if these surveys were restricted from this proposed MPA network.
%% TODO FIX THIS
%%
%%
%% The changes noted here may not meaningfully change most stock assessment outcomes, although this is a critical next question.
%% However, if spatial protection and associated restrictions for scientific surveys continue to expand, index integrity would further degrade and eventually compromise the ability to manage exploited populations.
%% % Therefore, we suggest allowing at least some survey effort in closed areas, particularly from less destructive gear; accelerating research on less-destructive survey methods; and ensuring adequate calibration between any historical and new survey methods to maintain index integrity.
%% Therefore, we suggest careful consideration when regulating scientific surveys within MPA boundaries to understand risk tradeoffs between the conservation objectives of the MPAs and the accuracy of indices, and other data products such as sampled ages of fish, needed for the assessment and management of exploited populations.

\vspace{5mm}
\noindent
Keywords:
groundfish,
index of abundance,
MPA,
spatiotemporal or spatio-temporal,
survey effort reduction,
scientific survey exclusion

@@

% <!-- Paragraph 1: scientific surveys are important and pervasive -->
Scientific surveys form the backbone of population status assessment for exploited marine species \citep[e.g.,][]{doubleday1981, hilbornwalters1992, gunderson1993}. One of the primary products of such surveys are indices of relative or absolute stock biomass or abundance through time (herein population or stock ``indices''). These indices are vital to most fisheries stock assessment \citep{hilbornwalters1992} and to several other applications such as the assessment of species at risk \citep{iucn2012}. Because of this importance, government agencies spend considerable effort to maintain random, stratified random, or fixed-station sampling designs as well as consistent gear and survey timing so that indices remain proportional to underlying trends in population biomass or abundance.

% <!-- Paragraph 2: globally MPAs are being proposed more and more -->
Globally, regions of the ocean are increasingly set aside as reserves or marine protected areas (MPAs) to protect ecosystems, species groups, fisheries, and culturally and socially valued areas from a range of anthropogenic threats. Motivated in part by the direction of the United Nations Convention on Biological Diversity to protect at least 10\% of marine areas by 2020 (since increased to 30\% by 2030), the area of MPAs worldwide has increased from 0.7\% to $\sim$8\% over the last 22 years \citep{WDPA2022}. Although the level of protection of individual MPAs varies from highly protected to multiuse \citep{hortaecosta2016, grorud-colvert2021}, it is clear that full- or high-protection MPAs (i.e., low or no extractive or destructive activities) have the best conservation outcomes \citep[e.g.,][]{lester2008, sciberras2015, edgar2014, zupan2018}. Commercial bottom trawling is therefore frequently restricted in MPAs. In Canada, for example, excluding industrial bottom trawling has been recommended as a ``Protection Standard'' for all of conservation areas that meet the IUCN definition of an MPA \citep{dfo2022standards}.

% <!-- Paragraph 3: restricted from fishing, but then the question becomes scientific surveys -->
While MPA regulations typically target commercial and recreational fishing activities, a focus on building more effective MPAs has also raised concerns about impacts on benthic habitat from scientific survey gear \citep{field2006, ursgroup2016, saarman2018, benoit2020national}. Although research surveys typically have much lower benthic footprints than commercial fleets---primarily because of their considerably smaller scale \citep{benoit2020national}---hook-and-line survey gear, and in particular trawl gear, damages benthic habitat \citep[e.g.,][]{collie2000a, kaiser2006, hiddink2017}. In Canada, scientific research surveys that pose a risk to conservation objectives may be excluded from MPAs and fisheries closures. In most cases, a permitting process determines if the research activities constitute an exception under the MPA regulations. For example, Rockfish Conservation Areas, a collection of over 160 areas in Pacific Canada aiming to protect sensitive rockfish populations, prohibit several modes of commercial and recreational and fishing as well as trawl or longline extractive research surveys \citep{thornborough2020}.

% <!-- # Paragraph 4: if excluded, this presents a number of possible issues -->
If scientific surveys are excluded from MPAs based on the gear used and negative impacts to benthic conservation objectives, there are several concerns related to loss of precision, inter-annual accuracy, and trend bias in indices derived from these surveys \citep{field2006, rideout2017, saarman2018, benoit2020national, benoit2020gulf, ices2023}.
A reduced number of survey sets is expected to reduce index precision according to statistical sampling theory \citep{cochran1977}, although survey effort may be redirected to the remaining survey domain.
Loss of survey coverage from portions of a population's distribution may also introduce bias if relative abundance inside and outside the closed areas changes because of fish movement or recovery within closed areas \citep[][]{benoit2020national}.
An analyst faced with a restricted survey must furthermore decide whether to ``shrink'' the survey domain by removing the restricted area \citep{rideout2017, benoit2020national} or extrapolate/interpolate from the surveyed region into the restricted area \citep{benoit2020national}.
This decision may have implications for precision, accuracy, and bias and therefore scientific advice to management.

% <!-- # Paragraph 5: what we do here -->
Here, we develop a general approach for evaluating the impacts of restricting scientific surveys on population indices through a retrospective analysis of survey data.
We apply this approach to a proposed MPA network off western Canada---the Northern Shelf Bioregion MPA Network \citep[][Fig.~\ref{fig:map}]{dfo2022networkactionplan}---as a case study.
We use spatiotemporal-model-based and traditional design-based methods to derive indices of relative biomass or abundance for four surveys (three using trawl gear and one using long-line gear) and \nSpp\ groundfish species and recalculate those indices as if scientific surveys had been restricted from areas proposed in the MPA network.
We evaluate changes in index precision, accuracy, trend bias, and power to detect decline, and evaluate correlates of these changes.

@@

We derived both design-based and spatiotemporal-model-based (herein ``model-based'') indices of abundance.
% In many regions, design-based estimators have traditionally been used \citep{schnute2000} REFs; however, geostatistical model-based indices are increasingly used in stock assessments \citep[e.g.,][]{thorson2015a} REFs.
Our results, however, focus on model-based indices because of a general move towards this approach, because the approach tends to have better precision than design-based indices \citep{shelton2014, thorson2015a}, and because model-based approaches are likely better suited to dealing with unanticipated changes to survey design \citep{ices2023}.

@@

We derived model-based indices of relative abundance or biomass using a geostatistical modelling approach with spatial and spatiotemporal Gaussian Markov random fields describing latent spatial and spatiotemporal processes \citep[e.g.,][]{shelton2014, thorson2015a, anderson2022}.
Depending on the species and survey, these models were configured as either delta models \citep{aitchison1955} with a logit link for a binomial model of encounter probability  and log link for a gamma model of positive catch weight \citep{schnute2003} or as \citet{tweedie1984} observation models with a log link for catch weight.
We included an offset term \citep[][p.~206]{mccullagh1989} for log area swept for the trawl models and log hook count for the long-line models.
Following a common model configuration for index standardization \citep[e.g.,][]{thorson2015a, thorson2019a}, our models included an independent intercept each year, a fixed spatial random field, and independent spatiotemporal random fields.
We reduced model complexity as needed based on diagnostics to reflect an analyst aiming for adequate model convergence and parsimony (Table~\ref{tab:model-configs}, Supporting Material).
We fit our models with the sdmTMB \textsf{R} package \citep{anderson2022}, which implements the models through TMB \citep{kristensen2016} with input matrices generated through \textsf{R}-INLA \citep{lindgren2015}.
Additional modelling details are available in the Supporting Material.

@@

We calculated a series of indices. The first two form the main analysis and the third is an additional option we focus on mainly in supplemental figures.

@@

\item Shrunk: estimate the index based on survey observations outside the proposed MPA network and a reduced survey domain that excludes the MPA network. We expect this to be the most likely outcome if survey effort is restricted although in the future some survey effort may be redirected over the shrunk survey domain.

\item Extrapolated: using observations outside the MPA network,  estimate the area-weighted index of abundance for the full status quo survey domain. This option could take advantage of nearby data, spatial correlation, and possible covariates to predict density within the MPA and retain the full survey domain.
\end{enumerate}

@@

\begin{enumerate}
  \item Precision: We calculated the annual coefficient of variation $\mathrm{CV}_{t,s}$ for year $t$ and survey~$s$ as $\sqrt{\exp(\mathrm{SE}_{t,s}^2) - 1}$, where $\mathrm{SE}$ is the standard error on the index as calculated with the generalized delta method in TMB \citep{kristensen2016} for the model-based index. We then calculated the proportional change in $\mathrm{CV}_{t,s}$ between each restricted index and the status quo index. This represents a standardized measure of precision loss.

    \item Accuracy: We calculated the annual relative error (RE) as $(\hat{\theta}_t - \theta_t) / \theta_t$, where $\hat{\theta}_t$ represents the index estimate from the restricted model in year $t$ and $\theta_t$ represents the index estimate from the status quo survey. We first centered $\hat{\theta}_t$ and $\theta_t$ by dividing them by their respective geometric means across years to avoid including consistent bias (an index from a shrunk survey domain should be lower than a full-survey domain index) and because surveys are often considered relative indices of abundance/biomass with an estimated catchability parameter that is estimated in stock assessments. We then calculated the absolute RE. The median absolute RE (MARE) represents a measure of accuracy loss---the expected magnitude of relative error in a typical year.

@@

We examined three potential predictors of changes in precision, accuracy, and trend bias:
(1) the CV of the status quo index, (2) the biomass or abundance proportion of a species inside the MPA network, and (3) trends in proportion of a species inside vs.\ outside the MPA.
For the first two, we fit generalized linear models (GLMs) with gamma observation error and log links with log-transformed predictor variables to identify patterns.
This implies that a proportional change in the predictor corresponds to a proportional change in the response.
For any species with slight proportional decreases in precision, we set these values to small positive values (0.01) for the purposes of this GLM because the log link requires positive data.
For the third, we compared the slope of relative error per decade against slope of density proportion within the MPA per decade for the most impacted survey(s) and fit a linear regression.
% We did not fit these models to evaluate statistical significance (noting that we ignore uncertainty in the predictor and response variables) but to characterize the relationships.
Because our analysis applies to proposed MPAs, the MPAs should not have had any causal effect on historical biomass or abundance.

@@

We performed a power analysis to identify the reduced ability to detect an exponential decline in population trend under the MPA-restricted scenario compared to the status quo.
We focused our analysis on species with at least 15\% of their biomass or abundance density within the MPA network and only for the surveys that were most impacted in precision and accuracy.

We used the spatiotemporal models fit to the full dataset (Supporting Material) and simulated new datasets with the fitted spatial random fields, new simulated spatiotemporal random fields, and observation error based on the fitted distribution.
In the simulation, we replaced the year factor effects with exponential declines in biomass or abundance density as a linear function in log link space.
For models simulated from a delta-gamma distribution, we held the detection probability constant.
We chose rates of decline that would be meaningful and that illustrated contrast between the two sampling scenarios: 50\% decline for species with average status quo CVs of $>$ 0.25 and 30\% decline for species with CVs $\le$ 0.25.

We then fit spatiotemporal models (omitting spatiotemporal random effects for simplicity, consistency, and estimation speed) with the same observation family and tested whether the upper 95\% confidence interval (CI) on the slope parameter representing the effect of year was less than zero.
We repeated the simulation and estimation 120 times to calculate power: the proportion of times that the estimation models correctly identified a negative population trend.

@@

We applied the above approach to a specific MPA network and set of surveys off the west coast of Canada.
The MPA network planning area, the Northern Shelf Bioregion (NSB), is approximately 102,000~km$^2$ and the proposed MPA network encompasses about 30\% of this area (30,493~km$^2$).
% TODO: this still correct?
The network process has been co-led by 15 partner First Nations, the province of British Columbia, and the Canadian federal government (Parks Canada, Environment and Climate Change Canada, and Fisheries and Oceans Canada). 
% CR: Parks Canada, EEEC and DFO? Might be worth mentioning, just because they represent a diversity of spaces (not just fisheries management).
% TODO: ER: is this correct?
Following a systematic conservation planning approach \citep{margules2000}, its design was guided by national and regional goals \citep{canada2011, canada2014} and centered on meeting network conservation objectives \citep{dfo2022networkactionplan}.
Of the six goals of the network, Goal 1 is of primary importance: ``protect and maintain marine biodiversity, ecological representation and special natural features.''
Additional goals and objectives include the protection of traditional use and cultural heritage, contributing to the protection of fisheries resources and their habitat, and providing opportunities for scientific research, education, and awareness \citep{canada2014}.
The proposed network is built off existing areas, where nearly 50\% of the draft scenario footprint already exists as some type of spatial protection (existing provincial or federal MPAs or fisheries closures in the form of Rockfish Conservation Areas) leaving $\sim$50\% of the network as new areas identified through the planning process \citep{dfo2022networkactionplan}.
The proposed NSB MPA network was announced by the planning partners in February 2023 with plans for iterative implementation; the first set of new MPAs are expected to be established by 2025.

The NSB MPA network proposes `Category 1' and `Category 2' zones with the main distinction that Category 1 zones are proposed for implementation by 2025 and Category 2 zones by 2030 \citep{dfo2022networkactionplan}. 
In our analysis, we assume both zones are restricted from bottom trawl and long-line research surveys. 
From the existing Gwaii Haanas National Marine Conservation Area Reserve and Haida Heritage Site, we excluded trawl surveys from all ``Strict Protection'' zones, while only excluding the long-line survey from ``Strict Protection'' zones that overlap with Rockfish Conservation Areas. We do not exclude survey activity from remaining existing MPA zones where possible changes to activities are uncertain at this time \citep{dfo2022networkactionplan}.

@@

We analyzed data from four research surveys including the three synoptic bottom trawl surveys \citep[][Fig.~\ref{fig:map}]{sinclair2003syn, anderson2019synopsis}: 
% Queen Charlotte Sound \citep[SYN QCS,][]{williams2018synqcs}, 
% Hecate Strait \citep[SYN HS,][]{wyeth2018synhs}, and 
% West Coast Haida Gwaii \citep[SYN WCHG,][]{williams2018synwchg}. 
Queen Charlotte Sound (SYN QCS), 
Hecate Strait (SYN HS), and 
West Coast Haida Gwaii (SYN WCHG). 
% We omitted two related surveys off the west coast of Vancouver Island because they are not affected by the proposed MPA network. 
The included surveys cover the continental shelf and upper slope off the northern part of the west coast of Canada (Fig.~\ref{fig:map}) and primarily target soft-bottom habitat. They have been conducted since 2003, are typically run in alternating years, and follow a random depth-stratified design with sampling units of 2 km $\times$ 2 km blocks with consistent gear and fishing protocols \citep{sinclair2003syn}. The unit of measurement for the synoptic trawl surveys is fish biomass per area swept.

@@

Based on a threshold of at least 5\% of survey sets detecting a species, we selected \nSpp\ species of benthic fishes including rockfish (\textit{Sebastes}), flatfish (\textit{Pleuronectiformes}), Pacific cod (\textit{Gadus macrocephalus}), and several elasmobranch species (Table~\ref{tab:spp-sci}). Of these species, \synNSpp\ met this threshold in the synoptic trawl surveys and \hbllNSpp\ met this threshold in the longline survey.

@@

% \end{spacing}

@@

>>>>

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1.15in]{geometry}
\usepackage{amsmath, amsfonts, amssymb}
\usepackage[dvipsnames]{xcolor}
\definecolor{niceblue}{HTML}{236899}
\newcommand{\rev}[1]{{\color{niceblue} #1}}
\newcommand{\R}[1]{\label{#1}\linelabel{#1}}
\newcommand{\lr}[1]{line~\lineref{#1}}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=black,
    filecolor=black,
    urlcolor=niceblue,
    citecolor=niceblue,
    linkbordercolor = white
}
\usepackage{longtable}
\usepackage{amsbsy}
\usepackage{epsfig}
\usepackage{bm}
\usepackage{xspace}
\usepackage{color}
\usepackage{lineno}
\usepackage{ragged2e}
\usepackage{comment}

@@

% \doublespacing
% \onehalfspacing
\begin{document}
\begin{spacing}{1.4}
\maketitle

% \RaggedRight

@@

% ICES: 200 words max
% perhaps could frame this as broader than MPAs (e.g., mostly wind farms in the US) and emphasize the general approach more with the NSB MPA case study

Marine protected areas (MPAs) are increasingly common worldwide, typically restricting fishing activities. However, MPAs may also limit scientific surveys that impact benthic habitat. We develop an approach that combines historical data degradation and simulation to investigate the effects of excluding surveys from MPAs on population abundance indices. Our approach quantifies losses in precision, inter-annual accuracy, trend accuracy, and the power to detect trends, as well as correlates of these effects. We apply this approach to a proposed MPA network off western Canada, examining \nSpp\ groundfish species observed by four surveys. Survey exclusion particularly impacted less precise indices, species well-represented in MPAs, and those whose density shifted in or out of MPAs. Redistributing survey effort outside MPAs improved precision but had limited benefits for accuracy or trend detection---it \R{B17}can make our estimates more precise about the ``wrong'' index. While these changes may not qualitatively alter stock assessment for many species, in some cases $\approx$30 percentage point reductions in the power to detect simulated 50\% population declines suggest meaningful impacts are possible. If survey restrictions continue expanding, index integrity could further degrade, eventually compromising the management of exploited populations. Regulating surveys within MPA boundaries therefore requires careful consideration to balance MPA objectives with the need for reliable data.

\vspace{5mm}
\noindent
Keywords:
groundfish,
index of abundance,
MPA,
spatiotemporal,
survey effort reduction,
scientific survey exclusion

@@

% <!-- Paragraph 1: scientific surveys are important and pervasive -->
Scientific surveys form the backbone of population status assessment for exploited marine species \citep{doubleday1981, hilbornwalters1992, gunderson1993}. One of the primary products of such surveys are indices of relative or absolute stock biomass or abundance through time (herein population or stock ``indices''). These indices are vital to most fisheries stock assessment \citep{hilbornwalters1992} and to other applications such as the assessment of species at risk \citep{iucn2012}. Because of this importance, government agencies spend considerable effort to maintain random, stratified random, or fixed-station sampling designs, as well as consistent gear and survey timing, so that indices remain proportional to underlying trends in population biomass or abundance.

% <!-- Paragraph 2: globally MPAs are being proposed more and more -->
Globally, regions of the ocean are increasingly set aside as reserves or marine protected areas (MPAs) to protect ecosystems, species groups, fisheries, and culturally and socially valued areas from a range of anthropogenic threats. Motivated in part by the direction of the United Nations Convention on Biological Diversity to protect at least 10\% of marine areas by 2020 (since increased to 30\% by 2030), the area of MPAs worldwide has increased from 0.7\% to $\sim$8\% over the last 22 years \citep{WDPA2022}. Although the level of protection of individual MPAs varies from highly protected to multiuse \citep{hortaecosta2016, grorud-colvert2021}, it is clear that full- or high-protection MPAs (i.e., low or no extractive or destructive activities) have the best conservation outcomes \citep[e.g.,][]{lester2008, sciberras2015, edgar2014, zupan2018}. Commercial bottom trawling is therefore frequently restricted in MPAs. In Canada, for example, excluding industrial bottom trawling has been recommended as a ``Protection Standard'' for all conservation areas that meet the IUCN definition of an MPA \citep{dfo2022standards}.

% <!-- Paragraph 3: restricted from fishing, but then the question becomes scientific surveys -->
While MPA regulations typically target commercial and recreational fishing activities, a focus on building more effective MPAs has also raised concerns about impacts on benthic habitat from scientific survey gear \citep{field2006, ursgroup2016, saarman2018, benoit2020national}. Although research surveys typically have much lower benthic footprints than commercial fleets---primarily because of their considerably smaller scale \citep{benoit2020national}---hook-and-line survey gear, and in particular trawl gear, damages benthic habitat \citep[e.g.,][]{collie2000a, kaiser2006, hiddink2017}. Scientific research surveys that pose a risk to conservation objectives may be excluded from MPAs and fisheries closures. In Canada, for example, a permitting process usually determines if research activities constitute an exception under MPA regulations. For example, Rockfish Conservation Areas, a collection of over 160 areas in Pacific Canada aiming to protect sensitive rockfish populations, prohibit several modes of commercial and recreational and fishing as well as trawl or longline extractive research surveys \citep{thornborough2020}.

% <!-- # Paragraph 4: if excluded, this presents a number of possible issues -->
If scientific surveys are excluded from MPAs based on the gear used and negative impacts to benthic conservation objectives, there are several concerns related to the precision, inter-annual accuracy, and bias in indices derived from these surveys \citep{field2006, rideout2017, saarman2018, benoit2020national, benoit2020gulf, ices2020, ices2023}.
A reduced number of survey sets is expected to reduce index precision according to statistical sampling theory \citep{cochran1977}, although survey effort may be redirected to the remaining survey domain.
If relative abundance inside and outside the closed areas changes because of fish movement or recovery within closed areas, a loss of survey coverage from portions of a population's distribution may also introduce bias \citep[][]{benoit2020national}.
An analyst faced with a restricted survey must furthermore decide whether to ``shrink'' the survey domain by removing the restricted area \citep{rideout2017, benoit2020national} or extrapolate/interpolate from the surveyed region into the restricted area \citep{benoit2020national}.
\R{A2}This decision can also impact index quality and therefore scientific advice to management (e.g., stock status and catch recommendations), for example by dissociating the survey domain from the stock domain in the former case or by imposing survey patterns in the restricted area based on surrounding observations in the latter case.

% <!-- # Paragraph 5: what we do here -->
Here, we develop a general approach for evaluating the impacts of restricting scientific surveys on population indices through a retrospective analysis of survey data.
We apply this approach to a proposed MPA network off western Canada---the Northern Shelf Bioregion MPA Network \citep[][Fig.~\ref{fig:map}]{dfo2022networkactionplan}.
We use spatiotemporal-model-based and traditional design-based methods to derive indices of relative biomass or abundance for four surveys (three using trawl gear and one using long-line gear) and \nSpp\ groundfish species and recalculate those indices as if scientific surveys had been restricted from areas proposed in the MPA network.
We evaluate changes in index precision, accuracy, trend bias, and power to detect decline, and evaluate correlates of these changes.

@@

We derived both design-based and spatiotemporal-model-based (herein ``model-based'') indices of abundance from survey data.
Our results, however, focus on model-based indices because of a general move towards this approach, because the approach tends to have better precision than design-based indices \citep{thorson2015a}, and because model-based approaches are likely better suited to dealing with unanticipated changes to survey design \citep{ices2023}.

@@

We derived model-based indices of relative abundance or biomass using a geostatistical modelling approach with spatial and spatiotemporal Gaussian Markov random fields describing latent spatial and spatiotemporal processes \citep[e.g.,][]{thorson2015a, anderson2022}.
Depending on the species and survey, these models were configured as either delta models \citep{aitchison1955} with a logit link for a binomial model of encounter probability  and log link for a gamma model of positive catch weight \citep{schnute2003} or as \citet{tweedie1984} observation models with a log link for catch weight.
We included an offset term \citep[][p.~206]{mccullagh1989} for log area swept for the trawl models and log hook count \R{B10} for the long-line models.
Following a common model configuration for index standardization \citep[e.g.,][]{thorson2015a, thorson2019a}, our models included an independent intercept each year, a fixed spatial random field, and independent spatiotemporal random fields.
We reduced model complexity as needed based on diagnostics to reflect an analyst aiming for adequate model convergence and parsimony (Table~\ref{tab:model-configs}, Supporting Material).
We fit our models with the sdmTMB \textsf{R} package \citep{anderson2022}, which implements the models through TMB \citep{kristensen2016} with input matrices generated through \textsf{R}-INLA \citep{lindgren2015}.
Additional modelling details are available in the Supporting Material.

@@

We calculated a series of indices. The first two form the main analysis and the third is an additional option we focus on in supplemental figures.

@@

\item Shrunk: estimate the index based on survey observations outside the proposed MPA network and a reduced survey domain that excludes the MPA network. We expect this to be the most likely outcome if survey effort is restricted.

\item Extrapolated: using observations outside the MPA network, estimate the area-weighted index of abundance for the full status quo survey domain. This option could take advantage of nearby data and spatial correlation to predict density within the MPA and retain the full survey domain.
\end{enumerate}

@@

\begin{enumerate}
    \item Precision: We calculated the annual coefficient of variation $\mathrm{CV}_{t}$ for year $t$ as $\sqrt{\exp(\mathrm{SE}_{t}^2) - 1}$, where $\mathrm{SE}$ is the standard error on the index as calculated with the generalized delta method in TMB \citep{kristensen2016} for the model-based index. We then calculated the proportional change in $\mathrm{CV}_{t}$ between each restricted index and the status quo index. This represents a standardized measure of precision loss.

    \item Accuracy: We calculated the annual relative error (RE) as $(\hat{\theta}_t - \theta_t) / \theta_t$, where $\hat{\theta}_t$ represents the index estimate from the restricted model in year $t$ and $\theta_t$ represents the index estimate from the status quo survey. We first centered $\hat{\theta}_t$ and $\theta_t$ by dividing them by their respective geometric means across years to avoid including consistent bias (an index from a shrunk survey domain should be lower than a full-survey domain index) and because surveys are often considered relative indices of abundance or biomass with an estimated catchability parameter that is estimated in stock assessments. We then calculated the absolute RE. The median absolute RE (MARE) represents a measure of accuracy loss---the expected magnitude of relative error in a typical year.

@@

We examined three potential predictors of changes in precision, accuracy, and trend bias:
(1) the CV of the status quo index, (2) the biomass or abundance proportion of a species inside the MPA network, and (3) trends in proportion of a species inside vs.\ outside the MPA.
When modelling MARE as the response, we fit generalized linear models (GLMs) with gamma observation error and log links with log-transformed predictor variables to identify patterns.
This implies that a proportional change in the predictor corresponds to a proportional change in the response.
% For any species with slight proportional decreases in precision, we set these values to small positive values (0.01) for the purposes of this GLM because the log link requires positive data.
When modelling percent increase in CV as the response (which had some negative values), we fit a model with both the mean and observation variance growing linearly with the log-transformed predictor (log proportion stock in MPA).
% TODO loss)!

TODO!!!!!!!!!!!!!!!!!!!!!!!!!!! FIX LOG PREDICTOR?? WHAT DID I DO !!!!????????????????

For the third, we compared the slope of relative error per decade against slope of density proportion within the MPA per decade for the most impacted survey(s) and fit a linear regression.
% We did not fit these models to evaluate statistical significance (noting that we ignore uncertainty in the predictor and response variables) but to characterize the relationships.
Because our analysis applies to proposed MPAs, the MPAs should not have had any causal effect on historical biomass or abundance.

@@

We performed a power analysis to identify the reduced ability to detect an exponential population decline under the MPA-restricted scenario compared to the status quo.
We used the spatiotemporal models fit to the full dataset (Supporting Material) and simulated new datasets with the fitted spatial random fields, new simulated spatiotemporal random fields, and observation error based on the fitted distribution.
In the simulation, we replaced the year factor effects with exponential declines in biomass or abundance density as a linear function in log link space.
For models simulated from a delta-gamma distribution, we held the detection probability constant.
We chose rates of decline that would be meaningful \R{A4}in a conservation context and that illustrated contrast between the two sampling scenarios: 50\% decline for species with average status quo CVs of $>$ 0.25 and 30\% decline for species with CVs $\le$ 0.25 over the span of the surveys (15 to 18 years).

We then fit spatiotemporal models (omitting spatiotemporal random effects for simplicity, consistency, and estimation speed) with the same observation family and tested whether the upper 95\% confidence interval (CI) on the slope parameter representing the effect of year was less than zero.
We repeated the simulation and estimation 300 times to calculate power: the proportion of times that the estimation models correctly identified a negative population trend.

@@

We applied the above approach to an MPA network and set of surveys off the west coast of Canada.
The MPA network planning area, the Northern Shelf Bioregion (NSB), is approximately 102,000~km$^2$ and the proposed MPA network encompasses about 30\% of this area (30,493~km$^2$).
% TODO: this still correct?
The network process has been co-led by 15 partner First Nations, the province of British Columbia, and the Canadian federal government.
% (Fisheries and Oceans Canada, Environment and Climate Change Canada, Parks Canada, Natural Resources Canada, Transport Canada).
% CR: Parks Canada, EEEC and DFO? Might be worth mentioning, just because they represent a diversity of spaces (not just fisheries management).
% TODO: ER: is this correct?
Following a systematic conservation planning approach \citep{margules2000}, its design was guided by national and regional goals \citep{canada2011, canada2014} and centered on meeting network conservation objectives \citep{dfo2022networkactionplan}.
Of the six goals of the network, Goal 1 is of primary importance: ``protect and maintain marine biodiversity, ecological representation and special natural features.''
Additional goals and objectives include the protection of traditional use and cultural heritage, contributing to the protection of fisheries resources and their habitat, and providing opportunities for scientific research, education, and awareness \citep{canada2014}.
The proposed network is built off existing areas, where nearly 50\% of the draft scenario footprint already exists as some type of spatial protection (existing provincial or federal MPAs or fisheries closures in the form of Rockfish Conservation Areas) leaving $\sim$50\% of the network as new areas identified through the planning process \citep{dfo2022networkactionplan}.
The proposed NSB MPA network was announced by the planning partners in February 2023 with plans for iterative implementation; the first set of new MPAs are expected to be established by 2025.
Our analysis makes some necessary assumptions about what areas of the network may be restricted from survey activity (Supporting Material: Northern Shelf Bioregion zones).

@@

We analyzed data from four research surveys including three synoptic bottom trawl surveys \citep[][Fig.~\ref{fig:map}]{sinclair2003syn, anderson2019synopsis}:
% Queen Charlotte Sound \citep[SYN QCS,][]{williams2018synqcs},
% Hecate Strait \citep[SYN HS,][]{wyeth2018synhs}, and
% West Coast Haida Gwaii \citep[SYN WCHG,][]{williams2018synwchg}.
Queen Charlotte Sound (SYN QCS),
Hecate Strait (SYN HS), and
West Coast Haida Gwaii (SYN WCHG).
% We omitted two related surveys off the west coast of Vancouver Island because they are not affected by the proposed MPA network.
The included surveys cover the continental shelf and upper slope off the northern part of the west coast of Canada (Fig.~\ref{fig:map}) and primarily target soft-bottom habitat. They have been conducted since 2003, are typically run in alternating years, and follow a random depth-stratified design with sampling units of 2 km $\times$ 2 km blocks with consistent gear and fishing protocols \citep{sinclair2003syn}. The unit of measurement for the synoptic trawl surveys is fish biomass per area swept.

@@

Based on a threshold of at least 5\% of survey sets detecting a species, we selected \nSpp\ species of benthic fishes including rockfishes (\textit{Sebastes}), flatfishes (\textit{Pleuronectiformes}), Pacific cod (\textit{Gadus macrocephalus}), and several elasmobranch species (Table~\ref{tab:spp-sci}). Of these species, \synNSpp\ met this threshold in the synoptic trawl surveys and \hbllNSpp\ met this threshold in the longline survey.

@@

\end{spacing}

@@

\subsection*{Northern Shelf Bioregion zones}
